Ionization of biological molecules by multicharged ions using the stoichiometric model

### 2019-11-13 ionmol_stoi.tex
change file names
JPB reviewers' comments

\begin{abstract}
In the present work, we investigate the ionization of molecules of 
biological interest by the impact of multicharged ions in the 
intermediate to high energy range. We performed full non--perturbative 
distorted--wave calculations (CDW) for thirty--six collisional systems 
composed by six atomic targets: H, C, N, O, F and S --which are the 
constituents of most of the DNA and biological molecules-- and six 
charged projectiles (antiprotons, H, He, B, C, and O). On account of 
the radiation damage caused by secondary electrons, we inspect the 
energy and angular distributions of the emitted electrons from the 
atomic targets. We examine seventeen molecules: DNA and RNA bases, 
DNA backbone, pyrimidines, tetrahydrofuran (THF), and C$_n$H$_n$ 
compounds. We show that the simple stoichiometric model (SSM), which 
approximates the molecular ionization cross sections as a linear 
combination of the atomic ones, gives reasonably good results for 
complex molecules. We also inspect the extensively used Toburen scaling 
of the total ionization cross sections of molecules with the number of 
weakly bound electrons. Based on the atomic CDW results, we 
propose new active electron numbers, which improves significantly 
scaling for all the targets and ions studied here in the intermediate 
to the high energy region. The new scaling describes well the available 
experimental data for proton impact, including small molecules. We 
perform full molecular calculations for five nucleobases and test a 
modified stoichiometric formula based on the Mulliken charge of the 
composite atoms. The difference introduced by the new stoichiometric 
formula is less than 3\%, which indicates the reliability of the SSM to 
deal with this type of molecules. The results of the extensive 
ion--target examination included in the present study allow us to assert 
that the SSM and the CDW--based scaling will be useful tools in this area.
\end{abstract}

@@

In the present study, we consider six atoms, $\alpha=$ H, C, N, O, P, 
and S, and six projectiles, antiprotons $\bar{p}$, H$^{+}$, He$^{2+}$, 
Be$^{4+}$, C$^{6+}$, and O$^{8+}$. 
Most of the organic molecules are composed of these atoms. Some 
particular molecules also include halogen atoms such as fluor and 
bromine; ionization cross sections of these elements have been 
previously published~\cite{miraglia2008}.

@@

\begin{figure*}[t!]
\centering
\includegraphics[width=0.8\textwidth]{atomicscaling.eps}
\caption{Reduced CDW total ionization cross section of six atomic 
targets. The curves are labeled with the charge state corresponding to 
the six multicharged projectiles.}
\label{fig:atomscaling}
\end{figure*}

@@

\begin{figure*}[t!]
\centering
\includegraphics[width=0.8\textwidth]{ener_mean.eps}
\caption{Mean emitted energy distribution for ionization by the impact 
of multicharged ions, given by Eq.~(\ref{40}). Dashed lines for $\bar{p}$,
solid lines for ion charges $+1$, $2+$, $4+$, $6+$ and $8+$,
as indicated.}
\label{fig:emittedener}
\end{figure*}

@@

\begin{figure*}[t!]
\centering
\includegraphics[width=0.8\textwidth]{ang_mean.eps}
\caption{Mean emitted angle distribution for ionization by impact of
multicharged ions.}
\label{fig:emittedang}
\end{figure*}

As mentioned before, secondary electrons contribute to the total damage. 
Then, not only the ejection energy is essential but also the angle 
of emission. Once again, we can consider the single differential cross 
section in terms of the ejected electron solid angle $\Omega$, 
$d\sigma_{\alpha,nl}/d\Omega$, to be expressed as a distribution function, 
and the mean angle $\overline{\theta}_{\alpha}$ can be defined as
\begin{equation}
\overline{\theta}_{\alpha}=\frac{\langle\theta_{\alpha}\rangle}
{\langle 1\rangle}=\frac{1}{\sigma_{\alpha}}\sum\limits_{nl}
\int d\Omega\,\theta\,\frac{d\sigma_{\alpha,nl}}{d\Omega}
\end{equation}

@@

\begin{figure*}[t!]
\centering
\includegraphics[width=0.8\textwidth]{adn1.eps}
\caption{Reduced CDW ionization cross section by the impact of 
multicharged ions. Experiments: \mbox{\Large$\circ$}~\cite{iriki2011} 
for proton impact and $\square$~\cite{rahman2016} for electron impact 
with equivelocity conversion.}
\label{fig:crossDNA_1}
\end{figure*}

\begin{figure*}[t!]
\centering
\includegraphics[width=0.8\textwidth]{adn2.eps}
\caption{Reduced CDW ionization cross section by the impact of 
multicharged ions. Experiments: proton impact on 
$\triangle$ uracil~\cite{itoh2013}, 
$\bigtriangledown$ pyrimidine~\cite{wolff2014} and $\meddiamond$
THF~\cite{wang2016}. Impact of $\ominus$ C$^{4+}$, 
$\oplus$ C$^{6+}$, O$^{6+}$, F$^{6+}$, and
$\otimes$ O$^{8+}$, F$^{8+}$ on 
uracil~\cite{agnihotri2012,agnihotri2013}. 
Symbols~$\rhd$~\cite{bug2017}, $\lhd$~\cite{wolf2019}, and 
$\medstar$~\cite{fuss2009} for electron impact with equivelocity 
conversion.}
\label{fig:crossDNA_2}
\end{figure*}

@@

\begin{figure*}[t!]
\centering
\includegraphics[width=0.8\textwidth]{molscaling85.eps}
\caption{Scaled ionization cross section per weakly bound electron using
(a)~the Toburen numbers $\nu_{\alpha}^T$, and (b) our proposed numbers
$\nu_{\alpha}^{\text{CDW}}$. Experiments: proton impact on 
\mbox{\Large$\circ$} adenine~\cite{iriki2011}, 
$\triangle$ uracil~\cite{itoh2013}, 
$\bigtriangledown$ pyrimidine~\cite{wolff2014} and $\meddiamond$ 
THF~\cite{wang2016}; electron impact on $\rhd$ pyrimidine~\cite{bug2017},
and $\lhd$, $\medstar$~\cite{wolf2019,fuss2009} THF.}
\label{fig:newscaling}
\end{figure*}

@@

\begin{table}[t]
\begin{center}
\begin{tabular}{|p{0.07\textwidth}p{0.02\textwidth}|
p{0.095\textwidth}p{0.02\textwidth}|p{0.095\textwidth}p{0.035\textwidth}|}
\hline
 Molecule & $n_e'$ &Molecule          & $n_e'$ & Molecule             & $n_e'$\\
\hline
 H$_2$    & 2  & C$_2$H$_7$N         & 19    & C$_4$H$_5$N$_3$O     & 37   \\
 H$_2$O   & 6  & C$_4$H$_8$O         & 28    & C$_5$H$_6$N$_2$O$_2$ & 42   \\
 NH$_3$   & 7  & C$_4$H$_4$N$_2$     & 28    & C$_5$H$_5$N$_5$      & 45   \\
 CH$_4$   & 8  & C$_6$H$_6$          & 30    & C$_5$H$_5$N$_5$O     & 49   \\
 CH$_5$N  & 13 & C$_4$H$_4$N$_2$O$_2$& 36    & C$_5$H$_{10}$O$_5$P  & 54.5 \\
 \hline
\end{tabular}
\caption{New scaling numbers for some molecular targets of biological 
interest.}
\label{nn}
\end{center}
\end{table}

\begin{figure*}[t!]
\centering
\includegraphics[width=0.785\textwidth]{scale_ne.eps}
\caption{Ionization cross sections by the impact of protons at 0.5, 1,
and 2 MeV in terms of the number of active electrons given by Table~\ref{nn}.
Experiments: 
\mbox{\Large$\circ$}~adenine~\cite{iriki2011}, 
$\triangle$ uracil~\cite{itoh2013}, 
$\bigtriangledown$ pyrimidine~\cite{wolff2014}, 
$\blacktriangle$ C2H$_7$N, CH$_5$N, methane and ammonia~\cite{lynch1976},
\mbox{\scriptsize$\bigstar$} ammonia and H$_2$~\cite{rudd1985}, and 
\mbox{\Large$\bullet$} water~\cite{luna2007}.}
\label{fig:recta}
\end{figure*}

@@

\begin{figure*}[t]
\centering
\includegraphics[width=0.8\textwidth]{levelsDNA.eps}
\caption{Theoretical molecular binding energies for adenine, cytosine, 
guanine, thymine, and uracil compared to those of atomic constituents.}
\label{fig:bindener}
\end{figure*}

@@

\begin{table*}[t]
\begin{center}
\begin{tabular}{|p{0.08\textwidth}|p{0.05\textwidth}|p{0.05\textwidth}|p{
0.05\textwidth}|p{0.05\textwidth}|p{0.19\textwidth}|}
\hline
Element & C & H & N & O & New stoichiometry \\
\hline
Adenine & +0.32 & +0.23 & --0.55 &       & 
C$_{4.92}$H$_{4.77}$N$_{5.14}$ \\ 
\hline
Cytosine & +0.28 & +0.21 & --0.56 & --0.53 & 
C$_{3.93}$H$_{2.79}$N$_{5.14}$O$_{1.13}$ \\ 
\hline
Guanine & +0.46 & +0.20 & --0.58 & --0.36 & 
C$_{4.89}$H$_{4.80}$N$_{5.15}$O$_{1.09}$ \\ 
\hline
Thymine & +0.20 & +0.19 & --0.54 & --0.52 & 
C$_{4.95}$H$_{1.81}$N$_{6.13}$O$_{2.13}$ \\ 
\hline
Uracil & +0.31 & +0.22 & --0.59 & --0.47 & 
C$_{3.92}$H$_{1.78}$N$_{4.15}$O$_{2.12}$ \\ 
\hline
\end{tabular}
\caption{Average effective Mulliken charge per atom $q_{\alpha}$, and 
new stoichiometric formula defined by Eq.~(\ref{eq:newstoi}) for five 
DNA molecules.}
\label{tab:newstoi}
\end{center}
\end{table*}

>>>>

\begin{abstract}
In the present work, we investigate the ionization of molecules of 
biological interest by the impact of multicharged ions in the 
intermediate to high energy range. We performed full non--perturbative 
distorted--wave calculations (CDW) for thirty--six collisional systems 
composed by six atomic targets: H, C, N, O, F and S --which are the 
constituents of most of the DNA and biological molecules-- and six 
charged projectiles (antiprotons, H, He, B, C, and O). On account of 
the radiation damage caused by secondary electrons, we inspect the 
energy and angular distributions of the emitted electrons from the 
atomic targets. We examine seventeen molecules: DNA and RNA bases, 
DNA backbone, pyrimidines, tetrahydrofuran (THF), and C$_n$H$_n$ 
compounds. We show that the simple stoichiometric model (SSM), which 
approximates the molecular ionization cross sections as a linear 
combination of the atomic ones, gives reasonably good results for 
complex molecules. We also inspect the extensively used Toburen scaling 
of the total ionization cross sections of molecules with the number of 
weakly bound electrons. Based on the atomic CDW results, we 
propose new active electron numbers, which leads to a more universal 
scaling for all the targets and ions studied here in the intermediate 
to the high energy region. The new scaling describes well the available 
experimental data for proton impact, including small molecules. We 
perform full molecular calculations for five nucleobases and test a 
modified stoichiometric formula based on the Mulliken charge of the 
composite atoms. The difference introduced by the new stoichiometric 
formula is less than 3\%, which indicates the reliability of the SSM to 
deal with this type of molecules. The results of the extensive 
ion--target examination included in the present study allow us to assert 
that the SSM and the CDW--based scaling will be useful tools in this area.
\end{abstract}

@@

In the present study, we consider six atoms, $\alpha=$ H, C, N, O, P, 
and S, and six projectiles, antiprotons $\bar{p}$, H$^{+}$, He$^{2+}$, 
Be$^{4+}$, C$^{6+}$, and O$^{8+}$. 
Most of the organic molecules are composed of these atoms. Some 
particular molecules also include halogen atoms such as fluorine and 
bromine; ionization cross sections of these elements have been 
previously published~\cite{miraglia2008}.

@@

\begin{figure*}[t!]
\centering
\includegraphics[width=0.8\textwidth]{figuras/atomicscaling.eps}
\caption{Reduced CDW total ionization cross section of six atomic 
targets. The curves are labeled with the charge state corresponding to 
the six multicharged projectiles.}
\label{fig:atomscaling}
\end{figure*}

@@

\begin{figure*}[t!]
\centering
\includegraphics[width=0.8\textwidth]{figuras/ener_mean.eps}
\caption{Mean emitted energy distribution for ionization by the impact 
of multicharged ions, given by Eq.~(\ref{40}). 
Solid lines for ion charges $+1$, $2+$, $4+$, $6+$ and $8+$, as indicated.
Dashed lines for $\bar{p}$ and dotted line for Born with $Z=1$.}
\label{fig:emittedener}
\end{figure*}

@@

\begin{figure*}[t!]
\centering
\includegraphics[width=0.8\textwidth]{figuras/ang_mean.eps}
\caption{Mean emitted angle distribution for ionization by impact of
multicharged ions. Curves as in Fig.~\ref{fig:emittedener}.}
\label{fig:emittedang}
\end{figure*}

As mentioned before, secondary electrons contribute to the total damage. 
Then, not only the ejection energy is essential but also the angle 
of emission. Once again, we can consider the single differential cross 
section in terms of the ejected electron solid angle $\Omega$, 
$d\sigma_{\alpha,nl}/d\Omega$, to be expressed as a distribution function, 
and the mean angle $\overline{\theta}_{\alpha}$ can be defined as
\begin{eqnarray}
\overline{\theta}_{\alpha}&=&\frac{\langle\theta_{\alpha}\rangle}
{\langle 1\rangle}=\frac{1}{\sigma_{\alpha}}\sum\limits_{nl}
\int d\Omega\,\theta\,\frac{d\sigma_{\alpha,nl}}{d\Omega} \\
\langle 1\rangle &=&\sigma_{\alpha}=\sum\limits_{nl}\int d\Omega\,
\frac{d\sigma_{\alpha,nl}}{d\Omega}\,, 
\end{eqnarray}

@@

\begin{figure*}[t!]
\centering
\includegraphics[width=0.8\textwidth]{figuras/adn1.eps}
\caption{Reduced CDW ionization cross section by the impact of 
multicharged ions. Experiments: \mbox{\Large$\circ$}~\cite{iriki2011} 
for proton impact and $\square$~\cite{rahman2016} for electron impact 
with equivelocity conversion.}
\label{fig:crossDNA_1}
\end{figure*}

\begin{figure*}[t!]
\centering
\includegraphics[width=0.8\textwidth]{figuras/adn2.eps}
\caption{Reduced CDW ionization cross section by the impact of 
multicharged ions. Experiments: proton impact on 
$\triangle$ uracil~\cite{itoh2013}, 
$\bigtriangledown$ pyrimidine~\cite{wolff2014} and $\meddiamond$
THF~\cite{wang2016}. Impact of $\ominus$ C$^{4+}$, 
$\oplus$ C$^{6+}$, O$^{6+}$, F$^{6+}$, and
$\otimes$ O$^{8+}$, F$^{8+}$ on 
uracil~\cite{agnihotri2012,agnihotri2013}. 
Symbols~$\rhd$~\cite{bug2017}, $\lhd$~\cite{wolf2019}, and 
$\medstar$~\cite{fuss2009} for electron impact with equivelocity 
conversion.}
\label{fig:crossDNA_2}
\end{figure*}

@@

\begin{figure*}[t!]
\centering
\includegraphics[width=0.8\textwidth]{figuras/molscaling85.eps}
\caption{Scaled ionization cross section per weakly bound electron using
(a)~the Toburen numbers $\nu_{\alpha}^T$, and (b) our proposed numbers
$\nu_{\alpha}^{\text{CDW}}$ for molecules listed in Table~\ref{tab:families}. 
For each band, the molecules are ordered from the smallest (top curve) to the
largest (bottom curve). Experiments: proton impact on 
\mbox{\Large$\circ$} adenine~\cite{iriki2011}, 
$\triangle$ uracil~\cite{itoh2013}, 
$\bigtriangledown$ pyrimidine~\cite{wolff2014} and $\meddiamond$ 
THF~\cite{wang2016}; electron impact on $\rhd$ pyrimidine~\cite{bug2017},
and $\lhd$, $\medstar$~\cite{wolf2019,fuss2009} THF.}
\label{fig:newscaling}
\end{figure*}

@@

\begin{table}[t]
\begin{center}
\begin{tabular}{|p{0.07\textwidth}p{0.02\textwidth}p{0.02\textwidth}|
p{0.095\textwidth}p{0.02\textwidth}p{0.02\textwidth}|
p{0.095\textwidth}p{0.035\textwidth}p{0.02\textwidth}|}
\hline
 Molecule & $n_e'$ & $n_e$ & Molecule          & $n_e'$ & $n_e$ & Molecule             & $n_e'$ & $n_e$ \\
\hline
 H$_2$    & 2 & 2   & C$_2$H$_7$N         & 19 & 20 & C$_4$H$_5$N$_3$O     & 37 & 42 \\
 H$_2$O   & 6 & 8   & C$_4$H$_8$O         & 28 & 30 & C$_5$H$_6$N$_2$O$_2$ & 42 & 48 \\
 NH$_3$   & 7 & 8   & C$_4$H$_4$N$_2$     & 28 & 30 & C$_5$H$_5$N$_5$      & 45 & 50 \\
 CH$_4$   & 8 & 8   & C$_6$H$_6$          & 30 & 30 & C$_5$H$_5$N$_5$O     & 49 & 56 \\
 CH$_5$N  & 13 & 14 & C$_4$H$_4$N$_2$O$_2$& 36 & 40 & C$_5$H$_{10}$O$_5$P  & 54.5 & 65 \\
 \hline
\end{tabular}
\caption{New scaling numbers, $n_e'$, and Toburen numbers, $n_e$,
for some molecular targets of biological interest.}
\label{nn}
\end{center}
\end{table}

\begin{figure*}[t!]
\centering
\includegraphics[width=0.785\textwidth]{figuras/scale_ne.eps}
\caption{Ionization cross sections by the impact of protons at 0.5, 1,
and 2 MeV in terms of the number of active electrons given by Table~\ref{nn}.
Experiments: 
\mbox{\Large$\circ$}~adenine~\cite{iriki2011}, 
$\triangle$ uracil~\cite{itoh2013}, 
$\bigtriangledown$ pyrimidine~\cite{wolff2014}, 
$\blacktriangle$ C$_2$H$_7$N, CH$_5$N, methane and ammonia~\cite{lynch1976},
\mbox{\scriptsize$\bigstar$} ammonia and H$_2$~\cite{rudd1985}, and 
\mbox{\Large$\bullet$} water~\cite{luna2007}.}
\label{fig:recta}
\end{figure*}

@@

\begin{figure*}[t]
\centering
\includegraphics[width=0.8\textwidth]{figuras/levelsDNA.eps}
\caption{Theoretical molecular binding energies for adenine, cytosine, 
guanine, thymine, and uracil compared to those of atomic constituents.}
\label{fig:bindener}
\end{figure*}

@@

\begin{table*}[t]
\begin{center}
\begin{tabular}{|p{0.08\textwidth}|p{0.05\textwidth}|p{0.05\textwidth}|p{
0.05\textwidth}|p{0.05\textwidth}|p{0.19\textwidth}|}
\hline
Element & C & H & N & O & New stoichiometry \\
\hline
Adenine & +0.32 & +0.23 & --0.55 &       & 
C$_{4.92}$H$_{4.77}$N$_{5.14}$ \\ 
\hline
Cytosine & +0.28 & +0.21 & --0.56 & --0.53 & 
C$_{3.93}$H$_{4.79}$N$_{3.14}$O$_{1.13}$ \\ 
\hline
Guanine & +0.46 & +0.20 & --0.58 & --0.36 & 
C$_{4.89}$H$_{4.80}$N$_{5.15}$O$_{1.09}$ \\ 
\hline
Thymine & +0.20 & +0.19 & --0.54 & --0.52 & 
C$_{4.95}$H$_{5.81}$N$_{2.13}$O$_{2.13}$ \\ 
\hline
Uracil & +0.31 & +0.22 & --0.59 & --0.47 & 
C$_{3.92}$H$_{3.78}$N$_{2.15}$O$_{2.12}$ \\ 
\hline
\end{tabular}
\caption{Average effective Mulliken charge per atom $q_{\alpha}$, and 
new stoichiometric formula defined by Eq.~(\ref{eq:newstoi}) for five 
DNA molecules.}
\label{tab:newstoi}
\end{center}
\end{table*}
